# NCAA Women's Volleyball Statistical Analysis Exercise: 2024 vs 2025

## Step 1: Load Required Libraries and Data


```{r}
library(tidyverse)
```


```{r}
matches24 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/main/data/ncaa_womens_volleyball_matchstats_2024.csv")

matches25 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/main/data/ncaa_womens_volleyball_matchstats_2025.csv")

```

## Step 2: Explore Maryland's Aces Performance

Let's examine how Maryland's serving (measured by aces) varied by match outcome across both seasons.

```{r}
# 2024 Maryland aces by match result
matches24 |>
  filter(team == 'Maryland') |>
  group_by(result) |>
  summarise(
    avg_aces = mean(aces),
    median_aces = median(aces),
    matches = n()
  )
```

```{r}
# 2025 Maryland aces by match result  
matches25 |>
  filter(team == 'Maryland') |>
  group_by(result) |>
  summarise(
    avg_aces = mean(aces, na.rm = TRUE),
    median_aces = median(aces, na.rm = TRUE),
    matches = n()
  )
```

**Question 1:** 
- How did their ace performance change between 2024 and 2025?
- Did Maryland serve more aces in wins or losses in each season? 
- What might explain any differences you observe? You can do some research for this.

YOUR ANSWERS HERE

## Step 3: Prepare Data for Statistical Testing

Let's test out whether there is a significant difference in average aces per match between the two seasons. First, we need to isolate Maryland's data and add a season identifier.

```{r}
umd_aces24 <- matches24 |> 
  filter(team == 'Maryland') |> 
  mutate(season = 2024)

umd_aces25 <- matches25 |> 
  filter(team == 'Maryland') |> 
  mutate(season = 2025)
```

Let's examine the distributions before testing:

```{r}
# Create side-by-side boxplots
combined_preview <- bind_rows(umd_aces24, umd_aces25)

ggplot(combined_preview, aes(x = factor(season), y = aces)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.6) +
  labs(title = "Maryland Aces Distribution by Season",
       x = "Season", y = "Aces per Match") +
  theme_minimal()
```

**Question 2:** Based on the boxplot, what do you notice about the distribution of aces in each season? Do you see any outliers?

YOUR ANSWER HERE

## Step 4: Conduct T-Test for Season Comparison

Let's test out whether there's a significant difference in average aces per match between the two seasons using a t-test.

**Null Hypothesis:** There is no difference in average aces per match between 2024 and 2025.
**Alternative Hypothesis:** There is a difference in average aces per match between the two seasons.

```{r}
result <- t.test(umd_aces24$aces, umd_aces25$aces, 
                alternative = "two.sided", 
                var.equal = TRUE)

print(result)
```

**Question 3:** 
- What is the p-value? Is it statistically significant at α = 0.05?
- What does the confidence interval tell you?
- Based on these results, can you reject the null hypothesis? What does this mean in practical terms?

YOUR ANSWERS HERE

## Step 5: Alternative Team Analysis

Let's repeat the analysis for a different team. UTEP had the highest number of aces in 2024 with 263. Has that success continued for the Miners this season?

```{r}
team_name <- "UTEP"

team_aces24 <- matches24 |> 
  filter(team == team_name) |> 
  mutate(season = 2024)

team_aces25 <- matches25 |> 
  filter(team == team_name) |> 
  mutate(season = 2025)

# Repeat the t-test
team_result <- t.test(team_aces24$aces, team_aces25$aces, 
                     alternative = "two.sided", var.equal = TRUE)
print(team_result)
```
**Question 4:** 
- What is the p-value? Is it statistically significant at α = 0.05?
- What do the sample means tell you?
- Based on these results, can you reject the null hypothesis? What does this mean in practical terms?

YOUR ANSWERS HERE


**Final Reflection Questions:**
1. How do the results for UTEP compare to Maryland's results?
2. What factors might explain differences in ace performance between seasons for volleyball teams?
3. What are the limitations of using aces as a measure of team performance?
4. What columns from the data could you add to your analysis to get a fuller picture of aces' impact?

YOUR ANSWERS HERE

